---
title: "Customizable Table Building with tangram.pipe"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{customizable-table-building-with-tangram.pipe}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


# Overview
The goal of this package is to iteratively build a customizable data table, one 
row at a time. This package will allow a user to input a data object, specify 
the rows and columns to use for the summary table, and select the type of data
to use for each individual row. Missing data, overall statistics, and
comparison tests can be calculated using this package as well.

# Installation
```{r, eval=FALSE}
install.packages("tangram.pipe")
```

# Getting Started

## Loading supplementary packages
```{r}
suppressPackageStartupMessages(require(tangram.pipe))
suppressPackageStartupMessages(require(knitr))
suppressPackageStartupMessages(require(kableExtra))
```

# Initializing the table
The first step to using this package is to initialize the data table to create.
Here, the user will select the name of the dataset to be analyzed in the table and
specify the variable name to use for the columns. In addition, the user will need
to determine whether to account for missing data, calculated overall statistics 
across all columns, or conduct comparison tests across the columns for each row. 
The arguments for `missing`, `overall`, and `comparison` will be used as the 
defaults for each subsequent row added to the table; however, a user can specify
a different entry for each argument for individual rows if desired.

This vignette will use the built-in `iris` dataset, which is a well-known dataset
containing flower measurements for three species of iris flowers.  Since most of 
the data in `iris` is numerical, we will add in three additional made-up variables
(flower color, leaf color, and stem size) in order to demonstrate table-building
functions for non-numeric data. Note that the additional columns are made-up 
purely for demonstration of this package.

```{r}
iris$color <- sample(c("Blue", "Purple"), size=150, replace=TRUE)
iris$Stem.Size <- sample(c("Small", "Medium", "Medium", "Large"), size=150, replace=TRUE)
iris$Leaf.Color <- "Green"
iris[149,5] <- NA
iris[150,c(1:4, 6:8)] <- NA
head(iris) %>% kable(escape=T, align="cl") %>% trimws %>% kable_styling(c("striped","bordered"))
```

For this example, the variable 'Species' will be chosen as the column variable;
`missing` and `comparison` will be set to `FALSE` to generate a simple example.
```{r}
info <- tbl_start(data=iris, col_var="Species", missing=FALSE, overall=TRUE, comparison=FALSE)
```

Using this function creates a list object that stores the user preferences for 
building the table going forward; in addition to the five elements listed here, 
the number of rows is also saved to the list. Subsequent entries to the list
will store information for the rows, which will ultimately be compiled to create
the final table after all row information has been added.

# Adding Rows
## Numeric Rows
To start off, we will first add a numeric row to the table. The function `num_row`
reads in data that is numeric in form, and by default calculates the five-number
summary statistics (minimum, first quartile, median, third quartile, maximum), as
well as the mean and standard deviation for the numeric variable within each
column.  Since we specified `overall=TRUE` in the initialization step, an overall
summary row will be included as well. The default summary function is `num_default`, 
but the user may write their own function to calculate different summary statistics
from what is shown here. Currently, `num_default` is the only built-in summary
function for this data type.

Let's start by calculating summary statistics for the Sepal Length in the `iris`
dataset. Since it makes more sense to display the variable name as "Sepal Length"
rather than the R-generated "Sepal.Length", we will use the `rowlabels` argument 
to make this change for the table.
```{r}
info <- info %>% num_row(row_var="Sepal.Length", rowlabels="Sepal Length")
info[[7]] %>% kable(escape=T, align="cl") %>% trimws %>% kable_styling(c("striped","bordered"))
```

By default, each row function will use two decimal places in reported statistics.
We can use the `digits` argument to specify more or fewer significant digits in
the reported table.
```{r, echo=FALSE}
info <- tbl_start(data=iris, col_var="Species", missing=FALSE, overall=TRUE, comparison=FALSE)
```
```{r}
info <- info %>% num_row(row_var="Sepal.Length", rowlabels="Sepal Length", digits=4)
info[[7]] %>% kable(escape=T, align="cl") %>% trimws %>% kable_styling(c("striped","bordered"))
```

There is a small amount of missing data within the `iris` dataset. Currently, 
`num_row` filters out the missing data and only considers data with complete cases
of the row and column variables. To see how much missing data there is in the
sepal length, we specify `missing=TRUE`.
```{r, echo=FALSE}
info <- tbl_start(data=iris, col_var="Species", missing=FALSE, overall=TRUE, comparison=FALSE)
```
```{r}
info <- info %>% num_row(row_var="Sepal.Length", rowlabels="Sepal Length", missing=TRUE)
info[[7]] %>% kable(escape=T, align="cl") %>% trimws %>% kable_styling(c("striped","bordered"))
```

The function above tells us that the dataset is missing a sepal length measurement
for three virginica flowers and one setosa flower, for a total of four missing
data entries. Note that the function cannot locate instances of missingness
in the column variable.

Finally, suppose we want to look at the differences in means across all species.
The function `num_diff` for the `comparison` argument will calculated the mean
difference in sepal length for each row compared to a reference category, which
is coded as the first column variable in the table. Here, versicolor and virginica
will be compared to setosa. The function also provides a 95% Confidence interval
to accompany the mean difference. Currently, `num_diff` is the only built-in comparison
function for `num_row`.
```{r, echo=FALSE}
info <- tbl_start(data=iris, col_var="Species", missing=FALSE, overall=TRUE, comparison=FALSE)
```
```{r}
info <- info %>% num_row(row_var="Sepal.Length", rowlabels="Sepal Length", comparison=num_diff)
info[[7]] %>% kable(escape=T, align="cl") %>% trimws %>% kable_styling(c("striped","bordered"))
```

## Categorical Rows
Now, we will look at adding categorical variables. The function `cat_row`
reads in data that is categorical in form, and by default calculates the number
of instances for each row category within each column category, as well as the 
column-wise proportions. The default summary function is `cat_default`, 
but the user may write their own function to calculate different summary statistics
from what is shown here. Currently, `cat_default` is the only built-in summary
function for this data type.

We will demonstrate this function by looking at `Stem.Size` in the `iris` dataset.
Note that all possible arguments for `cat_row` are the same as those for `num_row`
```{r}
info <- info %>% cat_row("Stem.Size", rowlabels="Stem Size")
info[[8]] %>% kable(escape=T, align="cl") %>% trimws %>% kable_styling(c("striped","bordered"))
```

Setting `missing=TRUE` will reveal the proportion of each species that does not
have a corresponding entry for stem size
```{r, echo=FALSE}
info <- tbl_start(data=iris, col_var="Species", missing=FALSE, overall=TRUE, comparison=FALSE)
info <- info %>% num_row(row_var="Sepal.Length", rowlabels="Sepal Length")
```
```{r}
info <- info %>% cat_row("Stem.Size", rowlabels="Stem Size", missing=TRUE)
info[[8]] %>% kable(escape=T, align="cl") %>% trimws %>% kable_styling(c("striped","bordered"))
```

Finally, let's look at a comparison test for a categorical row. The default
comparison function is `cat_comp_default`, which will calculate the relative
entropy between each column and the reference category, as well as conduct
a CBhi-Square Goodness of Fit test on the data present. Currently, 
`cat_comp_default` is the only built-in function for categorical data, but a 
user may write their own function to use instead.
```{r, echo=FALSE}
info <- tbl_start(data=iris, col_var="Species", missing=FALSE, overall=TRUE, comparison=FALSE)
info <- info %>% num_row(row_var="Sepal.Length", rowlabels="Sepal Length")
```
```{r}
info <- info %>% cat_row("Stem.Size", rowlabels="Stem Size", comparison=cat_comp_default)
info[[8]] %>% kable(escape=T, align="cl") %>% trimws %>% kable_styling(c("striped","bordered"))
```

## Binary Data
The final type of data we will examine here is binary data; this is when a variable
can only take on two possible values. In a table, it can be helpful to only include
one of the options if the second entry can be deduced from looking at the first.
This is done using the `binary_row` function. Summary statistics are calculated
using `binary_default`, and are the same as those calculated using a categorical
variable. Note that a user may use `cat_row` to process binary data if they wish
to see both row entries included in the table.

We will now demonstrate the use of `binary_row` on the color variable in `iris`.
In the dataset, the available colors are blue and purple, so we do not wish
to include both entries here.
```{r}
info <- info %>% binary_row("color", rowlabels="Color")
info[[9]] %>% kable(escape=T, align="cl") %>% trimws %>% kable_styling(c("striped","bordered"))
```

The `binary_row` function includes all of the same arguments as the previous row
functions, but additionally includes a new argument, `reference`.  This allows 
a user to choose which group will appear on the table. By default, the 
alphabetically first row group will appear on the table, which is why 'Blue' 
appeared above. If we want to see the statistics for purple flowers, we can
run the following code.
```{r, echo=FALSE}
info <- tbl_start(data=iris, col_var="Species", missing=FALSE, overall=TRUE, comparison=FALSE)
info <- info %>% num_row(row_var="Sepal.Length", rowlabels="Sepal Length")
info <- info %>% cat_row("Stem.Size", rowlabels="Stem Size")
```
```{r}
info <- info %>% binary_row("color", rowlabels="Color", reference="Purple")
info[[9]] %>% kable(escape=T, align="cl") %>% trimws %>% kable_styling(c("striped","bordered"))
```

Finally, let's look at some comparison functions used for binary data. By default,
this row function will calculate the difference in proportions by using `binary_diff`
if `comparison=TRUE` during initialization. This will calculated differences in
proportions across columns; the calculations will also include 95% Confidence intervals.
```{r, echo=FALSE}
info <- tbl_start(data=iris, col_var="Species", missing=FALSE, overall=TRUE, comparison=FALSE)
info <- info %>% num_row(row_var="Sepal.Length", rowlabels="Sepal Length")
info <- info %>% cat_row("Stem.Size", rowlabels="Stem Size")
```
```{r}
info <- info %>% binary_row("color", rowlabels="Color", comparison=binary_diff)
info[[9]] %>% kable(escape=T, align="cl") %>% trimws %>% kable_styling(c("striped","bordered"))
```

The package has two additional options for comparison tests using binary data. 
Odds ratios can be calculated using `binary_or`, and risk ratios can be calculated
with `binary_rr`. Note that if `comparison=TRUE` is initialized in `tbl_start` and 
a user wants to use an odds ratio or risk ratio here, `comparison` must be set to 
either of those two options in this row addition, since excluding the argument
will lead to `binary_diff` being called by default.
```{r, echo=FALSE}
info <- tbl_start(data=iris, col_var="Species", missing=FALSE, overall=TRUE, comparison=FALSE)
info <- info %>% num_row(row_var="Sepal.Length", rowlabels="Sepal Length")
info <- info %>% cat_row("Stem.Size", rowlabels="Stem Size")
```
```{r}
info <- info %>% binary_row("color", rowlabels="Color", comparison=binary_or)
info[[9]] %>% kable(escape=T, align="cl") %>% trimws %>% kable_styling(c("striped","bordered"))
```

```{r, echo=FALSE}
info <- tbl_start(data=iris, col_var="Species", missing=FALSE, overall=TRUE, comparison=FALSE)
info <- info %>% num_row(row_var="Sepal.Length", rowlabels="Sepal Length")
info <- info %>% cat_row("Stem.Size", rowlabels="Stem Size")
```
```{r}
info <- info %>% binary_row("color", rowlabels="Color", comparison=binary_rr)
info[[9]] %>% kable(escape=T, align="cl") %>% trimws %>% kable_styling(c("striped","bordered"))
```

## Adding an empty row
The `empty_row` function will add a blank row to the final table. This is useful
if a user wants to include blank space between some of table's rows. The user only
needs to specify the name of the list object in order to create the blank row.
An optional argument is a header to include, should the user want to create a label
for the subsequent rows that follow in the table.
```{r}
info <- info %>% empty_row()
```

# Creating a Finished Product
The following code will generate a finalized table for the `iris` dataset.  It
will include all four numeric variables (sepal length, sepal width, petal length,
petal width), as well as stem size and color. The final table itself is generated
using `tbl_out`; a `print` command will display the final result.
```{r, eval=FALSE}
info <- tbl_start(iris, "Species", missing=FALSE, overall=TRUE, comparison=FALSE) %>%
  num_row("Sepal.Length", rowlabels="Sepal Length") %>%
  empty_row() %>%
  num_row("Sepal.Width", rowlabels="Sepal Width") %>%
  empty_row() %>%
  num_row("Petal.Length", rowlabels="Petal Length") %>%
  empty_row() %>%
  num_row("Petal.Width", rowlabels="Petal Width") %>%
  empty_row() %>%
  cat_row("Stem.Size", rowlabels="Stem Size") %>%
  empty_row() %>%
  binary_row("color", rowlabels="Color") %>%
  tbl_out() %>% 
  print()
```
```{r, echo=FALSE}
info <- tbl_start(iris, "Species", missing=FALSE, overall=TRUE, comparison=FALSE) %>%
  num_row("Sepal.Length", rowlabels="Sepal Length") %>%
  empty_row() %>%
  num_row("Sepal.Width", rowlabels="Sepal Width") %>%
  empty_row() %>%
  num_row("Petal.Length", rowlabels="Petal Length") %>%
  empty_row() %>%
  num_row("Petal.Width", rowlabels="Petal Width") %>%
  empty_row() %>%
  cat_row("Stem.Size", rowlabels="Stem Size") %>%
  empty_row() %>%
  binary_row("color", rowlabels="Color") %>%
  tbl_out()
  info[[18]] %>% kable(escape=T, align="cl") %>% trimws %>% kable_styling(c("striped","bordered"))
```

Setting `comparison=TRUE` will compute the default comparison functions for each
row in the table. Rows that do not correspond to the comparison type will be
left blank in that column.
```{r, eval=FALSE}
info <- tbl_start(iris, "Species", missing=FALSE, overall=TRUE, comparison=TRUE) %>%
  num_row("Sepal.Length", rowlabels="Sepal Length") %>%
  empty_row() %>%
  num_row("Sepal.Width", rowlabels="Sepal Width") %>%
  empty_row() %>%
  num_row("Petal.Length", rowlabels="Petal Length") %>%
  empty_row() %>%
  num_row("Petal.Width", rowlabels="Petal Width") %>%
  empty_row() %>%
  cat_row("Stem.Size", rowlabels="Stem Size") %>%
  empty_row() %>%
  binary_row("color", rowlabels="Color") %>%
  tbl_out() %>% 
  print()
```
```{r, echo=FALSE}
info <- tbl_start(iris, "Species", missing=FALSE, overall=TRUE, comparison=TRUE) %>%
  num_row("Sepal.Length", rowlabels="Sepal Length") %>%
  empty_row() %>%
  num_row("Sepal.Width", rowlabels="Sepal Width") %>%
  empty_row() %>%
  num_row("Petal.Length", rowlabels="Petal Length") %>%
  empty_row() %>%
  num_row("Petal.Width", rowlabels="Petal Width") %>%
  empty_row() %>%
  cat_row("Stem.Size", rowlabels="Stem Size") %>%
  empty_row() %>%
  binary_row("color", rowlabels="Color") %>%
  tbl_out()
  info[[18]] %>% kable(escape=T, align="cl") %>% trimws %>% kable_styling(c("striped","bordered"))
```

# Additional Features
## Single Column of Data
The package can handle cases where there is only a single column of data. In the
`iris` dataset, if we set the column variable to be Leaf.Color, we see that there
is only one possible leaf color in the dataset. In this case, the column variable
statistics would be exactly the same as those for the overall column in the table.
```{r, eval=FALSE}
info <- tbl_start(iris, "Leaf.Color", missing=FALSE, overall=TRUE, comparison=FALSE) %>%
  num_row("Sepal.Length", rowlabels="Sepal Length") %>%
  tbl_out() %>%
  print()
```
```{r, echo=FALSE}
info <- tbl_start(iris, "Leaf.Color", missing=FALSE, overall=TRUE, comparison=FALSE) %>%
  num_row("Sepal.Length", rowlabels="Sepal Length") %>%
  tbl_out()
  info[[8]] %>% kable(escape=T, align="cl") %>% trimws %>% kable_styling(c("striped","bordered"))
```

# Changing datasets within a row
This package allows for an individual row to use a different dataset from the one
initialized in `tbl_start`. Use the `newdata` argument to specify the new dataset
to use, then define the rows and columns for the new data. Note that if a new row
is added after the row with the differing dataset, the new row will automatically
return to using the initialized dataset from `tbl_start` unless the user specifies
otherwise in `newdata`.

For this example, we will be using a new dataset called `birds`. This dataset
includes three species of bird, their corresponding ages, and the region
where they live.
```{r, echo=FALSE}
birds <- data.frame(Type=character(20), Age=numeric(20), Region=character(20))
type <- c("Cardinal", "Bluejay", "Owl")
birds$Type <- sample(type, 20, replace=TRUE)
birds$Age <- rnorm(20,10,1)
region=c("East", "West")
birds$Region <- sample(region, 20, replace=TRUE)
```
```{r}
head(birds) %>% kable(escape=T, align="cl") %>% trimws %>% kable_styling(c("striped","bordered"))
```

```{r, eval=FALSE}
info <- tbl_start(iris, "Species", missing=FALSE, overall=TRUE, comparison=FALSE) %>%
  num_row("Sepal.Length", rowlabels="Sepal Length") %>%
  empty_row() %>%
  num_row("Sepal.Width", rowlabels="Sepal Width") %>%
  empty_row(header="Switch to Birds Dataset") %>% 
  num_row(row_var="Age", col_var="Type", newdata=birds) %>%
  empty_row() %>% 
  cat_row(row_var="Region", col_var="Type", newdata=birds) %>%
  tbl_out() %>%
  print()
```
```{r, echo=FALSE}
info <- tbl_start(iris, "Species", missing=FALSE, overall=TRUE, comparison=FALSE) %>%
  num_row("Sepal.Length", rowlabels="Sepal Length") %>%
  empty_row() %>%
  num_row("Sepal.Width", rowlabels="Sepal Width") %>%
  empty_row(header="Switch to Birds Dataset") %>% 
  num_row(row_var="Age", col_var="Type", newdata=birds) %>%
  empty_row() %>% 
  cat_row(row_var="Region", col_var="Type", newdata=birds) %>%
  tbl_out()
  info[[14]] %>% kable(escape=T, align="cl") %>% trimws %>% kable_styling(c("striped","bordered"))
```

Since the columns of the rows with the `birds` data differ from the columns of `iris`, 
all columns are included in the table, but only columns corresponding to the data
used in the rows will have values filled in.










